---
title: "Biostat M280 Homework 4"
subtitle: Due Mar 16 @ 11:59PM
author: "Landi Luo (UID: 305027205)"
date: "3/9/2018"
output: 
  html_document:
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message = FALSE}
if (!require("pacman"))  
  install.packages("pacman", repos = "http://cran.us.r-project.org/")
p_load("sparklyr", "tidyverse", "ggmap", "maps", "mapdata", "knitr")
```

Our Apache Yarn cluster hosts the [flights](http://stat-computing.org/dataexpo/2009/the-data.html) data representing 123 million flights over 22 years. Read the [lecture notes](http://hua-zhou.github.io/teaching/biostatm280-2018winter/slides/12-sparklyr/sparklyr-flights.html) on how to access the Yarn cluster. Connect to the database using `sparklyr` and answer following questions. You can base your answers on a specific year or the whole data set.  

```{r, results="hide"}
# connect to spark
Sys.setenv(SPARK_HOME="/usr/lib/spark")
config <- spark_config()
sc <- spark_connect(master = "local", config = config)
sc
```
*Create dplyr reference to the Spark DataFrame.*   
```{r}
flights_tbl <- tbl(sc, 'flights')
airlines_tbl <- tbl(sc, 'airlines')
airports_tbl <- tbl(sc, 'airports')
    
flights_tbl
airlines_tbl
airports_tbl
```

1. Map the top 10 busiest airports. Size of dots should reflect the number of flights through that destination.  
Hint: You may find this tutorial on [Making Maps in R](http://eriqande.github.io/rep-res-web/lectures/making-maps-with-R.html) helpful.  
    *I mapped the top 10 busiest airports from the year 2002 by counting the total flights that took-off and landed from each airport. The busiest airport was Chicago O'Hare International Airport (ORD), which had 654,782 total flights.*  
    ```{r}
    # subset flights from 2002
    flights_yr <- flights_tbl %>%
      filter(year == 2002)
    
    flights_orig <- flights_yr %>% 
      # remove first row
      filter(origin != "Origin") %>%
      select(year, "airport" = origin ) 
    flights_dest <- flights_yr %>%
      filter(dest != "Dest") %>%
      select(year, "airport" = dest) 
    combine_orig <- flights_orig %>%  
      group_by(airport) %>%
      summarise(n = n()) %>%
      collect()
    combine_dest <- flights_dest %>%    
      group_by(airport) %>%
      summarise(n = n()) %>%
      collect()
    combined <- bind_rows(combine_orig, combine_dest) %>%
      group_by(airport) %>%
      summarise(`Total Flights` = sum(n)) %>%
      arrange(desc(`Total Flights`)) %>%
      head(10)
    
    # create vector with top 10 airport names
    target <- combined$airport
    
    airports2 <- airports_tbl %>%
      select(faa, lat, lon, name) %>%
      filter(faa %in% target) %>%
      collect()
    q1 <- left_join(combined, airports2, by = c("airport" = "faa"))
    # write_csv(q1, "finalq1.csv")
    
    ## generate map ##
     
    # convert lat and lon to numeric
    q1$lat <- as.numeric(q1$lat)
    q1$lon <- as.numeric(q1$lon)
    
    map <- get_googlemap(location = 'united states', zoom = 4)
    top_airports <- ggmap(map) + 
      geom_point(aes(x = lon, y = lat, size = `Total Flights`), 
                 data = q1, alpha = .5) +
      labs(x = "Longitude", y = "Latitude", 
           title = "Top 10 Busiest Airports in 2002")
    top_airports
    kable(combined, col.names = c("Airport", "Total Flights"), align = "l")
    ```

2. Map the top 10 busiest direct routes. Size of lines should reflect the number of flights through that route.  
    *I mapped the top 10 busiest direct routes for the year 2002.*
    ```{r}
    # get top 10 busiest routes
    routes <- flights_yr %>%
      filter(origin != "Origin") %>% 
      select(year, origin, dest) %>%
      group_by(origin, dest) %>%
      summarise(n = n()) %>%
      collect() %>%
      arrange(desc(n)) %>%
      head(10) %>%
      ungroup()
    
    # get latitude and longitude for origin and dest
    orig_vector <- routes$origin
    airports_orig <- airports_tbl %>%
      select(faa, lat, lon, name) %>%
      filter(faa %in% orig_vector) %>%
      collect()
    location_orig <- left_join(routes, airports_orig, by = c("origin" = "faa")) %>%
      select(n, origin, lat_orig = lat, lon_orig = lon)
    
    dest_vector <- routes$dest
    airports_dest <- airports_tbl %>%
      select(faa, lat, lon, name) %>%
      filter(faa %in% dest_vector) %>%
      collect()
    location_dest <- left_join(routes, airports_dest, by = c("dest" = "faa")) %>%
      select(dest, lat_dest = lat, lon_dest = lon)
    
    # combine into one dataframe
    q2 <- bind_cols(location_orig, location_dest)
    # write_rds(q2, "q2.rds")
    q2
    
    ```


3. LAX:

    (a). Reproduce above plot. Visualize and explain some prominent features you observe. For example, what happened at points 1-5?

    (b). Visualize and explain seasonal effects.
  
    (c). Visualize and explain weekly effects.
  
    (d). Map top 10 destinations from LAX. Size of dots should reflect the number of flights from LAX to that destination.
    
4. Build a predictive model for the arrival delay (`arrdelay`) of flights flying from LAX. Use the same filtering criteria as in the [lecture notes](http://hua-zhou.github.io/teaching/biostatm280-2018winter/slides/12-sparklyr/sparklyr-flights.html) to construct training and validation sets. You are allowed to use a maximum of 5 predictors. The prediction performance of your model on the validation data set will be an important factor for grading this question.
    
5. Visualize and explain any other information you want to explore.

```{r}
spark_disconnect(sc)
```

  
