---
title: "Biostat M280 Homework 2"
subtitle: Due Feb 16 @ 11:59PM
author: "Landi Luo"
date: "2/12/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Q1

Read [Chapter 7](http://r4ds.had.co.nz/exploratory-data-analysis.html) (Exploratory Data Analysis) of _R for Data Science_ and do exercises 7.3.4, 7.4.1, 7.5.1.1, 7.5.2.1, and 7.5.3.1.

### Exercise 7.3.4

1. Explore the distribution of each of the `x`, `y`, and `z` variables in `diamonds`. What do you learn? Think about a diamond and how you might decide which dimension is the length, width, and depth.
    x variable has minimum value = 0 and maximum value = 10.74.
    y variable has range 0 to 58.9.
    z variable has range 0 to 31.8. 
    ```{r}
    ggplot(data = diamonds) + 
      geom_histogram(mapping = aes(x = x), binwidth = 0.5)  
    diamonds %>%
      count(min(x), max(x))
    
    ggplot(data = diamonds) + 
      geom_histogram(mapping = aes(x = y), binwidth = 0.5)
    diamonds %>%
      count(min(y), max(y))
    
    ggplot(data = diamonds) + 
      geom_histogram(mapping = aes(x = z), binwidth = 0.5)
    diamonds %>%
      count(min(z), max(z))
    
    ```


## Q2 (optional)

Read [Chapter 23](http://r4ds.had.co.nz/model-basics.html) (Model Basics) and [Chapter 24](http://r4ds.had.co.nz/model-building.html) (Model Building) of _R for Data Science_ and do exercises 24.2.3 and 24.3.5.

## Q3

Redo HW1 Q2 using tidyverse.

## Q4 (optional)

Redo HW1 Q3 on Hoffman2, except now we want to submit each `runSum.R` job to a different node in the cluster.
